---
title: "ENVS567-Beetles"
author: "Kelsey Graves"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup 
```{r, message = FALSE, warning = FALSE}
library(easystats)
library(data.table)
library(readxl)
library(ggplot2)
library(reshape2)
library(car)
library(tidyverse)
library(marginaleffects)
library(GGally)
library(DHARMa)
library(MASS)
library(lme4)
library(glmmTMB)
library(modelbased)
```

Question: 
What is the effect of temperature on the weight of beetles accounting for variation in species and sex? 

Explore Data: 
```{r}
beetle <- read.table("../data_raw/beetles.txt", header=TRUE)
```

```{r}
ggplot(data=beetle, aes(x=Temp.C, y=Dry.Mass.mg)) + geom_point() + facet_wrap(~Sex)

ggplot(data=beetle, aes(x=Temp.C, y=Dry.Mass.mg, color=Species)) + geom_boxplot()

ggplot(data=beetle, aes(x=Temp.C, y=Dry.Mass.mg, color=Species)) + geom_boxplot() + facet_wrap(~Sex)

ggplot(data=beetle, aes(x=Temp.C, y=Dry.Mass.mg, color=Sex)) + geom_boxplot() + facet_wrap(~Species)

ggplot(data=beetle, aes(x=Species, y=Dry.Mass.mg, color=Sex)) + geom_point()

ggplot(data=beetle, aes(x= Temp.C, y=Dry.Mass.mg, color=Species))+ geom_point() +geom_smooth(method="lm", se=F)
```

```{r}
beetles.lmm <- lmer(data=beetle, log(Dry.Mass.mg) ~ Temp.C + (1|Species) + (1|Sex))

beetles.lmm2 <- lmer(data=beetle, log(Dry.Mass.mg) ~ Temp.C * Species + (1|Sex))
```

```{r}
#random slopes
beetles.rslopes1 <- lmer(log(Dry.Mass.mg) ~ Temp.C + (Temp.C|Sex) + (Temp.C|Species), data = beetle)

beetles.rslopes2 <- lmer(log(Dry.Mass.mg) ~ Temp.C + (Temp.C|Species:Sex), data = beetle)

beetles.rslopes3 <- lmer(
  log(Dry.Mass.mg) ~ Temp.C + (Temp.C | Species) + (1 | Sex),
  data = beetle)

summary(beetles.rslopes1)
summary(beetles.rslopes2)
summary(beetles.rslopes3)
```

```{r}
compare_performance(beetles.lmm, beetles.lmm2, beetles.rslopes1, beetles.rslopes2, beetles.rslopes3)
```

```{r}
check_model(beetles.lmm2, check=c("linearity", "homogeneity", "normality"))
```

```{r}
beetles.lmm2.ml <- lmer(data=beetle, log(Dry.Mass.mg) ~ Temp.C * Species + (1|Sex), REML=FALSE)
summary(beetles.lmm2.ml)
```

```{r}
drop1(beetles.lmm2.ml, test="Chisq")
```

```{r}
r2(beetles.lmm2.ml)
```

```{r}
pred.mean <- estimate_relation(beetles.lmm2)

pred.rand <- estimate_relation(beetles.lmm2, include_random=TRUE)


ggplot(beetle, aes(Temp.C, log(Dry.Mass.mg), color = Species)) +
  geom_point(aes(shape = Sex)) +
  geom_smooth(aes(group = 1), method = "lm", color = "black", se = FALSE) +
  
  theme_classic()
```

Results:

We found a significant interaction between Temperature and Species (p < 0.001), indicating that mass is dependent on both temperature and species. The coefficient of Temp.C (-0.013) indicates that as temperature increases, there is a slight decrease in mass. There are large differences between the baseline mass of the reference species (AB = e^1.788 = 5.977) and the other beetle species. The fixed effects of Temperature and Species account for 99.4% of variability in mass. Including the random effect Sex, the model accounts for 99.7% of variability in mass.



