---
editor_options: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------

title: "PhoebeTeskey_Problems_ENVS567" author: "Phoebe Teskey" date:
"2026-01-22" output: html_document

```{r}
library(ggplot2)
library(reshape2)

```

# R Language

*Problem 1*

1.  Under highly favorable conditions, populations grow exponentially.
    However resources will eventually limit population growth and
    exponential growth cannot continue indefinitely. This phenomenon is
    described by the logistic growth function:

N_t=(KN_0)/〖N_0+(K-N_0)e〗\^(-rt)

where Nt is the population size at time t, N0 is the initial population
size, r is the rate of growth or reproductive rate, and K is the
carrying capacity.

Write a logistic growth function in R that also generates a plot with
time on the x axis and Nt on the y axis. Using that function create 3
separate plots for 20 days assuming an initial population size of 10
individuals and a carrying capacity of 1,000 individuals under 3 growth
rate scenarios (0.5, 0.8, 0.4). Plot the previous results (that is, 3
growth rate scenarios) in a single plot using ggplot2.

```{r}

logistic_growth <- function(t, N0, r, K, plot = TRUE) {
  Nt <- (K * N0) / (N0 + (K - N0) * exp(-r * t))
  
  if (plot) {
    plot(
      t, Nt,
      type = "l",
      lwd = 2,
      xlab = "Time (days)",
      ylab = "Population size (Nt)",
      main = paste("Logistic Growth (r =", r, ")")
    )
  }
  
  return(data.frame(time = t, Nt = Nt, r = r))
}

# Parameters
time <- 0:20
N0 <- 10
K <- 1000

# Growth rate scenarios
r1 <- 0.5
r2 <- 0.8
r3 <- 0.4

# Generate plots
out1 <- logistic_growth(time, N0, r1, K)
out2 <- logistic_growth(time, N0, r2, K)
out3 <- logistic_growth(time, N0, r3, K)

# Combine results into one data frame
all_results <- rbind(out1, out2, out3)

# Convert r to a factor for nicer legend labels
all_results$r <- factor(all_results$r)

# Plot
ggplot(all_results, aes(x = time, y = Nt, color = r)) +
  geom_line(linewidth = 1.2) +
  labs(
    x = "Time (days)",
    y = "Population size (Nt)",
    color = "Growth rate (r)",
    title = "Logistic Growth Under Different Growth Rates"
  ) +
  theme_minimal()
```

*Problem 2*

a)  A function (sum_n) that for any given value, say n, computes the sum
    of the integers from 1 to n (inclusive). Use the function to
    determine the sum of integers from 1 to 5,000.

```{r}
n <- 1:5000

sum_n <- function(n){
  sum <- sum(1:length(n))
  return(sum)
}

sum_n(n)

```

b)  A function (sqrt_round) that takes a number x as input, takes the
    square root, rounds it to the nearest whole number and then returns
    the result.

```{r}
sqrt_round <- function(x) {
  sqrt <- x^(1/2)
  rounded <- round(sqrt)
  return(rounded)
}

sqrt_round(13)
```

# Review of Statistics

*Problem 1*: The data in the file DesertBirdAbundance.csv are from a
survey of the breeding birds of Organ Pipe Cactus National Monument in
southern Arizona.

Draw a histogram of the data.

```{r}
birds <- read.csv("../data_raw/DesertBirdAbundance.csv")
str(birds)
birds_plot <- ggplot(birds, aes(x = abundance))+
  geom_histogram()
plot(birds_plot)
```

Calculate the median and the mean. In this particular case, which do you
think is the best measure of center, the mean or the median? Why?

```{r}
mean(birds$abundance)
median(birds$abundance)
```

In this case, the median is a more accurate representation of the data
because we have outliers like the white-winged dove that was seen over
600 times.

*Problem 2*: Normal human body temperature is 98.6 ºF. Researchers
obtained body-temperature measurements on randomly chosen healthy
people: 98.4, 98.6, 97.8, 98.8, 97.9, 99.0, 98.2, 98.8, 98.8, 99.0,
98.0, 99.2, 99.5, 99.4, 98.4, 99.1, 98.4, 97.6, 97.4, 97.5, 97.5, 98.8,
98.6, 100.0, 98.4.

a)  Draw a histogram of the data.

```{r}
healthy_temps <- c(98.4, 98.6, 97.8, 98.8, 97.9, 99.0, 98.2, 98.8, 98.8, 99.0, 98.0, 99.2, 99.5, 99.4, 98.4, 99.1, 98.4, 97.6, 97.4, 97.5, 97.5, 98.8, 98.6, 100.0, 98.4)

hist(healthy_temps)
```

b)  Are these measurements consistent with a population mean of 98.6 ºF?

```{r}
temp_test <- t.test(healthy_temps, mu = 98.6)
print(temp_test)
```

There is not a significant difference between the mean of healthy_temps
and the population mean of 98.6 The confidence interval indicates that
the true population mean lies between 98.2 and 98.8. Therefore, yes,
these measurements are consistent with the population mean of 98.6

*Problem 3*: The brown recluse spider often lives in houses throughout
central North America. A diet-preference study gave each of 41 spiders a
choice between two crickets, one live and one dead. 31 of the 41 spiders
chose the dead cricket over the live one. Does this represent evidence
for a diet preference?

```{r}
binom.test(31, 41, p = 0.5)
```

Yes, this is a significant result.

*Problem 4:* A bee biologist is analyzing whether there was an
association between bee colony number and type of forest habitat. We
expect that there is no habitat preference for bee colony number. Make a
barplot of the data and assess if no habitat preference is true based on
this data?

```{r}
tree <- c("oak", "hickory", "maple", "red cedar", "poplar")
bee_colonies <- c(33, 30, 29, 4, 4)
tree_bee <- data.frame(x = tree, y = bee_colonies)

chisq_result <- chisq.test(bee_colonies)
print(chisq_result)

ggplot(tree_bee, aes(x = tree, y = bee_colonies))+
  geom_col()
```

There clearly seems to be a habitat preference for oak, maple and
hickory trees. The chi-sqared test confirms that this distribution is
not normal (p-value \< 0.05).

*Problem 5:* You are hired by the US Department of Agriculture to
develop effective control practices for invasive plants
(data_ANOVA.xlsx). In particular, you are asked to investigate pesticide
controls for kudzu which is an invasive vine that grows in thick mats
that smother underlying plants. Two of the most widely used pesticides
for kudzu are glyphosate and triclopyr. To determine the effectiveness
of a single application of pesticide, you conduct an experiment in 18
plots that each had 50% kudzu cover. In mid-summer, you applied equal
amounts of 2% glyphosate to 6 plots, 2% triclopyr to 6 plots, and water
without pesticides to 6 plots. Then you returned in autumn and measured
the percent cover of kudzu in the plots.

a)  Transform the data from wide to long format.

```{r}
kudzu <- read.csv("../data_raw/data_ANOVA.csv")
kudzu_long <- melt(kudzu)
```

b)  Make a boxplot and perform an ANOVA.

```{r}
ggplot(kudzu_long, aes(x = variable, y = value))+
  geom_boxplot()
kudzu_anova <- aov(value~variable, kudzu_long)
summary(kudzu_anova)
```

This indicates a significant result.

*Problem 6:* Testosterone is known to predict aggressive behavior and is
involved in face shape during puberty. Researchers measured the face
width-to-height ratio of 21 university hockey players with the average
number of penalty minutes awarded per game for aggressive infractions.
The data are available in face.txt.

a)  Calculate the correlation coefficient.

```{r}
aggression <- read.table("../data_raw/face.txt", header = T, sep = "\t")
cor(x = aggression$Face, y = aggression$Penalty, method = "pearson")
```

b)  Does face shape predict aggression? Write the result in the form of
    an equation for the line.

```{r}
aggression_prediction <- lm(Penalty ~ Face, data = aggression)
summary(aggression_prediction)

```

y = 3.189x - 4.505

Face shape is significant in influencing aggression, but just barely (p
value = 0.04)

c)  What is the variation explained?

The variation explained is represented by the R\^2 value, which is .29.
This means that about 29% of the variation in penalty minutes can be
explained by the face width-to-height ratio.

d)  Make a scatterplot of the data and include the linear regression
    line in the plot.

```{r}
ggplot(aggression, aes(x = Face, y = Penalty))+
  geom_point()+
  geom_smooth(method = "lm", se = T)
```

# Linear and Non-Linear Models

```{r}
library(easystats)
library(DHARMa)
library(MASS)
library(lme4)
library(ecostats)
library(GGally)
library(car)
```

*Question 1:* Data in growth.txt come from a farm-scale trial of animal
diets. There are two factors: diet and supplement. Diet is a factor with
three levels: barley, oats and wheat. Supplement is a factor with four
levels: agrimore, control, supergain and supersupp. The response
variable is weight gain after 6 weeks. Analyze the data following this
structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4.
Validate model, 5. Test hypothesis and assess explanatory power, and 6.
Communication of results (plot + description of main results).

Question: What diet and supplement increases weight in mice after 6
weeks?

```{r}
#Exploratory analyses

growth <- read.table("../data_raw/growth.txt", header = T)
weight_gain_mice <- lm(gain ~ diet * supplement, data = growth)
plot(parameters(weight_gain_mice))

ggplot(data = growth, aes(x = diet, y = gain, color = supplement))+
  geom_boxplot()

```

```{r}
#Fit model

model3 <- lm(gain ~ diet * supplement, data = growth)
summary(model3)

```

```{r}
#Validate model

check_model(model3, check=c("linearity", "homogeneity", "normality"))
check_homogeneity(model3)
plot(check_homogeneity(model3))

```

```{r}
#Test hypothesis and assess explanatory power
anova(model3)

```

```{r}
#r^2
r2(model3)

```

Communication of results (plot + description of main results):

No significant interaction between diet and supplement. The effect of
the supplements was consistent across diets. Type of diet was
significant, (F2,36 = 83.52, p \< 0.001) and the barley diet had a
higher gain than barley or wheat. The supplement was also significant,
(F3,36 = 17.8, p \< 0.001). The model accounted for 82% of the variance
in weight gain, (R2 = 0.82) with standard error at 1.31.

```{r}
model3.means <- estimate_means(model3)
amount.means <- estimate_means(model3, by = "diet")

ggplot(data = model3.means, aes(x = diet, y = Mean, color=supplement)) +
  geom_pointrange(aes(ymin = CI_low, ymax = CI_high), position=position_dodge(width=0.5), size=0.5) +
  geom_pointrange(data=amount.means, aes(ymin = CI_low, ymax = CI_high), position=position_dodge(width=0.5), size=1.5, color="black") +
  geom_jitter(data = growth, aes(x=diet, y=gain, color=supplement), alpha=0.6, size=1.5,
              position=position_jitterdodge(jitter.width = 0.2))+
  scale_color_manual(values = c("firebrick4", "dodgerblue3", "green", "orange"))+
  ylab("Weight (g)")+
  xlab(NULL)+
  theme_classic()
```

*Question 2:* Respiratory rate (Y) is expected to depend on body mass
(X) by the power law, Y=aX, where  is the scaling exponent. The data
are available in respiratoryrate_bodymass.txt.

a)  Analyze the data by transforming the data and then fit a linear
    model.

```{r}
breathe <- read.table("../data_raw/respiratoryrate_bodymass.txt", header = T)
breathe$logBodyMass <- log(breathe$BodyMass)
breathe$logRespiration <- log(breathe$Respiration)

fit.log <- lm(logRespiration ~ logBodyMass, data = breathe)
summary(fit.log)

```

For every 1% increase in body mass, the respiratory rate increases by
approximately 0.86%

```{r}
ggplot(data = breathe, aes(x=Respiration, y=BodyMass))+
  geom_point()+
  geom_smooth(method ="lm", se = T, color = "red")+
  scale_y_continuous(trans = "log")+
  theme_classic()
```

```{r}
ggplot(data = breathe, aes(x=Respiration, y=BodyMass))+
  geom_point()+
  geom_smooth(color= "red", method = "glm", method.args = list(family = gaussian(link = 'log')))+
  theme_classic()
```

*Question 3:* Use the diminish.txt data (xv is explanatory
(independent), yv is response variable (dependent)) to perform a simple
linear regression and a polynomial (second-degree) regression.

a)  Compare both models and decide which model seems to be better?

b)  Make a scatterplot and include both regression lines.

```{r}
diminish <- read.table("../data_raw/diminish.txt", header = T)

fit.lm <- lm(yv ~ xv, data=diminish)

fit.poly <- lm(yv ~ xv + I(xv^2), data=diminish)
summary(fit.poly)
```

```{r}
anova(fit.lm, fit.poly, test='LRT')
```

```{r}
compare_performance(fit.lm, fit.poly)
```

```{r}
ggplot(data=diminish, aes(x=xv, y=yv))+
  geom_point()+
  geom_smooth(method = "lm", se = T, color = "red")+
  geom_smooth(method = "lm", formula = y ~ poly(x,2), se = T)+
  theme_classic()
```

*Question 4* Use the ozone.txt data to model the ozone concentration as
a linear additive (no interactions) function of wind speed, air
temperature and the intensity of solar radiation.

a)  Make a pairwise scatterplot and assess the collinearity of the
    explanatory variables

```{r}
ozone <- read.table("../data_raw/ozone.txt", header = T)

ggpairs(ozone)
```

```{r}
ozone.lm <- lm(ozone ~ rad+temp+wind, data=ozone)
summary(ozone.lm)
```

```{r}
check_model(ozone.lm, check=c("linearity", "homogeneity", "normality"))
```

```{r}
check_collinearity(ozone.lm)
```

```{r}
plot(check_collinearity(ozone.lm))
```

```{r}
anova(ozone.lm)
```

```{r}
anova(lm(ozone ~ wind+rad+temp, data=ozone))
```

```{r}
anova(lm(ozone ~ wind+temp+rad, data=ozone))
```

```{r}
Anova(ozone.lm, type = 2)
```

```{r}
r2(ozone.lm)
```

We explored the linear relationships between Ozone, Radiation, Wind and
Temperature. Temperature and Radiation variables were positively
associated with higher ozone levels (p-values \< 0.05). Overall, our
multiple regression model explained 60% (adjusted R2) of the variability
in Ozone levels.

```{r}
plot(parameters(ozone.lm))
```

```{r}
plot(parameters(ozone.lm, standardize = "refit"))
```

*Question 5:* Use data(CO2) and assess the effects of the variables
Treatment (categorical) and conc (ambient concentration, numerical) on
the response uptake. Analyze the data following this structure: 1.
Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5.
Test hypothesis and assess explanatory power, and 6. Communication of
results (plot + description of main results).

Question: Does ambient concentration of CO2 and treatment type affect
uptake of CO2 in plants?

```{r}
#Exploratory analyses

data(CO2)

ggplot(CO2, aes(x = uptake, y = conc, color = Treatment)) +
  geom_point()

```

```{r}
#Fit model

model4 <- lm(uptake ~ Treatment * conc, data = CO2)
summary(model4)

```

```{r}
#Validate model

check_model(model4, check=c("linearity", "homogeneity", "normality"))


```

```{r}
#Test hypothesis and assess explanatory power

anova(model4)

```

```{r}
Anova(model4)
```

```{r}
r2(model4)
```

Communicate Results: There was not enough evidence of different
relationships between concentration of CO2 and uptake of CO2 among
treatment types (F1,80 = 0.4, p-value = 0.5297). The chilled treatment
type have on average 5 less units of CO2 than the nonchilled treatment
type (F1,80 = 74.1267, p-value \< 0.001). There was a positive
relationship between concentration of CO2 and uptake of CO2 (b = 0.0198,
F1, 80 = 28.5535, p value \< 0.001). Overall, the model explained 31% of
the variance in uptake of CO2.

```{r}
#Communicate results with a plot
ggplot(CO2, aes(y=uptake, x=conc, color=Treatment))+
  geom_point()+
  geom_smooth(method = "lm", se = T)+
  scale_color_manual(values = c("green", "pink"))+
  ylab("Uptake of CO2")+
  xlab("Concentration of CO2")+
  theme_classic()
```

*Question 6*: Use the jaws.txt data (age explanatory, bone response).

```         
Perform a non-linear regression assuming an asymptotic exponential relationship: y=a(1-e^(-cx))

Perform a non-linear regression assuming a Michaelis-Menten model: y=ax/(1+bx)

Estimate the percentage of variation explained by both models (comparing them with a null model with only a constant).

Compare both models using AIC.

Make a scatterplot of the data and include both regression lines.
```

```{r}
jaws <- read.table("../data_raw/jaws.txt", header = T)

jaws.asymp<-nls(bone~a*(1-exp(-c * age)), data=jaws,
                start = list(a = 120, c = 0.05))
summary(jaws.asymp)
```

```{r}
summary(nls(bone ~ SSmicmen(age, a, b), data=jaws))
```

```{r}
rss<-13.9^2*51 # residual sum of squares

null<-lm(bone~1, data=jaws) # null model with only a constant
summary(null)
```

```{r}
tss<-31.01^2*52 # total sum of squares

100*(tss-rss)/tss
```

```{r}
fit_mm <- nls(bone ~ a * age / (1 + b * age), 
              data = jaws, 
              start = list(a = 20, b = 0.1))

aic_asymp <- AIC(jaws.asymp)
aic_mm <- AIC(fit_mm)

summary(aic_asymp)
summary(aic_mm)
```

```{r}


age_range <- data.frame(age = seq(min(jaws$age), max(jaws$age), length.out = 200))

age_range$asymp_pred <- predict(jaws.asymp, newdata = age_range)
age_range$mm_pred <- predict(fit_mm, newdata = age_range)

ggplot(jaws, aes(x = age, y = bone)) +
  geom_point(alpha = 0.6, color = "darkgrey") +
  # Add Asymptotic Exponential line
  geom_line(data = age_range, aes(y = asymp_pred, color = "Asymptotic Exponential"), size = 1) +
  # Add Michaelis-Menten line
  geom_line(data = age_range, aes(y = mm_pred, color = "Michaelis-Menten"), size = 1, linetype = "dashed") +
  # Formatting
  scale_color_manual(name = "Models", values = c("Asymptotic Exponential" = "red", "Michaelis-Menten" = "blue")) +
  labs(title = "Bone Growth vs. Age: Model Comparison",
       x = "Age",
       y = "Bone") +
  theme_minimal()
```

# GLMs

```{r}
library(glmmTMB)
```

*Problem 1* The data in stork.txt display the stress-induced
corticosterone levels circulating in the blood of European white storks
and their survival over the subsequent five years of study. Analyze the
data following this structure: 1. Question, 2. Exploratory analyses, 3.
Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory
power, and 6. Communication of results (plot + description of main
results).

Question: Does corticosterone level predict survival of European white
storks over a period of 5 years?

```{r}
#Exploratory analyses
stork <- read.table("../data_raw/stork.txt", header = T)

ggplot(stork, aes(x = Corticosterone, y = Survival))+
  geom_point()


```

Fit Model:

```{r}
stork.glm<-glm(Survival ~ Corticosterone, family=binomial(link="logit"), data = stork)
summary(stork.glm)
```

Validate Model:

```{r}
simulateResiduals(stork.glm, plot = T)
```

Test hypothesis and assess explanatory power:

```{r}
anova(stork.glm, test = "LRT")
```

```{r}
r2(stork.glm)
```

Communicate results: Animals with lower corticosterone levels have a
higher rate of survival (-0.07980, Chisq = 3.8382, p-value 0.0501). The
model explained 10% of the variance.

```{r}
ggplot(aes(x=Corticosterone, y=Survival), data=stork)+
  geom_point(size = 2, alpha = 0.9)+
  geom_smooth(method=glm, method.args = list(family = "binomial"), se=T)+
  theme_classic()
```

*Problem 2* The clusters.txt dataset contains the response variable
Cancers (the number of reported cancer cases per clinic) and the
explanatory variable Distance (the distance from a nuclear plant to the
clinic in kilometers). Analyze the data following this structure: 1.
Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5.
Test hypothesis and assess explanatory power, and 6. Communication of
results (plot + description of main results).

1.) Question: Do cancer cases occur more often closer to a nuclear power
plant?

```{r}
#2.) Exploratory analyses

cancer <- read.table("../data_raw/clusters.txt", header = T)

ggplot(data = cancer, aes(x = Cancers))+
  geom_histogram(color = "white", fill = "black")

```

3.) Fit model

```{r}
cancer.glm <- glm(Cancers ~ Distance, data = cancer, family = poisson)
summary(cancer.glm)
```

Validate Model:

```{r}
simulateResiduals(cancer.glm, plot = T)
```

```{r}
testDispersion(cancer.glm, plot = T)
```

```{r}

cancer.nb <- glm.nb(Cancers ~ Distance, data=cancer)
summary(cancer.nb)
```

Test hypotheses and assess explanatory power:

```{r}
simulateResiduals(cancer.nb, plot = T)
```

```{r}
testDispersion(cancer.nb, plot = T)
```

```{r}
anova(cancer.nb, test = "LRT")
```

```{r}
r2(cancer.nb)
```

Communicate Results:

There is no significant relationship between number of cancer cases and
distance from a nuclear power plant. (Chisq = 1.6741, p value 0.19). The
model explained 2.7% of the variance. Based on these results, we cannot
conclude that living closer to a power plant increases the number of
reported cancer cases.

```{r}
ggplot(data = cancer, aes(x = Distance, y = Cancers)) +
  geom_point()+
  geom_smooth(method = "glm", method.args = list(family = "poisson"), color = "red") +
  labs(x =  "Distance from nuclear power plant", y = "Cancer cases count")+
  theme_classic()
```

*Question 3* Use data(cbpp) from library(lme4) to model the number of
cases of cattle plague (incidence) as a function of period. Control for
herd size in your model. Analyze the data following this structure: 1.
Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5.
Test hypothesis and assess explanatory power, and 6. Communication of
results (plot + description of main results).

Question: Is the number of cases of cattle plague influenced by time
period or herd size?

```{r}
#Exploratory analyses
plague <- cbpp

ggplot(plague, aes(x = incidence, y = size))+
  geom_point()

ggplot(plague, aes(x = incidence, y = period))+
  geom_point()

```

Fit Model

```{r}
plague_glmm <- glmer(incidence ~ period + (1 | herd) + offset(log(size)), 
                    data = plague, 
                    family = poisson)
summary(plague_glmm)
```

Validate Model

```{r}
simulateResiduals(plague_glmm, plot = T)
```

Test hypothesis and assess explanatory power:

```{r}
drop1(plague_glmm, test = "Chisq")
```

```{r}
r2(plague_glmm)
```

Communicate Results: A GLMM with Poisson distribution log and an offset
for herd size revealed that period had a signigicant effect on plague
incidence (chisq = 22.51, df = 3, p\<0.001).

```{r}
pred <- estimate_relation(plague_glmm, at = "period", offset = 0)
pred.farm <- estimate_relation(plague_glmm, at = c("period", "herd"), offset = 0

ggplot(pred.farm, aes(x = period, y = Predicted, group = herd)) +
  geom_line(color = "darkgrey", alpha = 0.4) +
  geom_point(data = plague, aes(x = period, y = incidence / size, color = period),
             size = 1.5, alpha = 0.6, position = position_jitter(width = 0.1))+
  geom_line(data = pred, aes(x = period, y = Predicted, group = 1),
            color = "black", linewidth = 1.5)+
  geom_ribbon(data = pred, aes(x = period, ymin = CI_low, ymax = CI_high,
                               fill= "grey"), alpha = 0.2) +
  scale_color_manual(values=c("firebrick4", "dodgerblue3", "green", "pink"))+
  labs(title = "Plague Incidence Rate by Period", x = "Period",
       y = "Incidence of Plague rate")+
  theme_classic()
```

*Question 4* Use insects.csv to model the number of insects (count) as
an additive function (no interaction) of habitat (categorical) and
flowers (numerical). Previous studies have reported that habitat quality
tends to drive the detection of insects (some habitats have no insects
regardless of flowers). Analyze the data following this structure: 1.
Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5.
Test hypothesis and assess explanatory power, and 6. Communication of
results (plot + description of main results).

Question: Is the number of insects counted correlated with habitat type
and number of flowers present?

```{r}
#Exploratory analyses
insects <- read.csv("../data_raw/insects.csv", header = T)

insects$flowers <- as.factor(insects$flowers)

ggplot(insects, aes(x = flowers, y = count)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~habitat)
```

Fit Model

```{r}
insect.model <- glm(count ~ habitat + flowers,
             family = poisson(link = "log"),
             data = insects)

summary(insect.model)
```

Validate Model

```{r}
simulateResiduals(insect.model, plot = T)
```

```{r}
testDispersion(insect.model, plot = T)
```

```{r}
insect.nb <- glm.nb(count ~ habitat + flowers, data = insects)
summary(insect.nb)
```

```{r}
simulateResiduals(insect.nb, plot = T)
```

```{r}
testDispersion(insect.nb, plot = T)
```

```{r}
insect_nbzi<-glmmTMB(count ~ habitat + flowers, ziformula = ~habitat, family=nbinom1, data=insects)
summary(insect_nbzi)
```

```{r}
testDispersion(insect_nbzi, plot = T)
```

```{r}
simulateResiduals(insect_nbzi, plot = T)
```

Test hypotheses and assess explanatory power:

```{r}
drop1(insect_nbzi, test = "Chisq")
```

```{r}
Anova(insect_nbzi, type = 2, test = "Chisq")
```

```{r}
r2(insect_nbzi)
```

Communicate Results: Insect abundance varied significantly as a function
of both habitat type and flower abundance. Habitat had a significant
effect on insect counts (p \< 0.001). Flower abundance was also a
significant predictor of insect count (p \< 0.001). The model explained
96% of the variability in insect count.

```{r}

pred <- estimate_relation(insect_nbzi, at = c("flowers", "habitat"), length = 100)


ggplot(pred, aes(x = flowers, y = Predicted)) +  
    geom_line(aes(group = 1), color = "blue", linewidth = 1) +
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high, group = 1), alpha = 0.2, fill = "blue") +

  geom_point(data = insects,
             aes(x = flowers, y = count),
             color = "black",
             alpha = 0.6,
             position = position_jitter(width = 0.2, height = 0.2)) +
  
  facet_wrap(~habitat) +
  
  ylab("Number of insects") +
  xlab("Number of flowers") +
  
  theme_classic()
```

# Mixed Models

```{r}
library(performance)
library(emmeans)
library(dplyr)
```

*Problem 1* Researchers at the University of Arizona want to assess the
germination rate of saguaros using a factorial design, with 3 levels of
soil type (remnant, cultivated and restored) and 2 levels of
sterilization (yes or no). The same experimental design was deployed in
4 different greenhouses. Each of the unique treatments was replicated in
5 pots. 6 seeds were planted in each pot.

a)  How many fixed treatments (unique combinations) exist?

3 \* 2 \* 4 = 24 fixed treatments exist.

b)  What is the total number of pots?

24 \* 5 = 120 pots

c)  What is the total number of plants measured?

120 \* 6 = 720 plants measured

d)  Write the R code that you would use to analyze these data.

Columns would be named: Germination yes/no, soil type, sterilization
yes/no, greenhouse, pot #, and seedling \#

Dataframe would be called saguaro.

I would use this model: model \<- glmer(Germination \~ Soil \*
Sterilization + (1\|Greenhouse/Pot), data = saguaro_data, family =
binomial)

I would then simulate residuals and use this test: drop1(model, test =
"Chisq) r2(model)

*Problem 2* Use Estuaries.csv to model the counts of invertebrate
species (Total) between modified or pristine estuaries (Mod) including
Estuary as random effects (only intercepts). Then fit the same model but
now also including Site in the random effects. Select between both
models parsimoniously. Communication your results (plot + description of
main results).

Results: Adding Site as a random effect nested within Estuary improved
the model fit (p \< 0.001, chi\^2 = 60.7).

Modification significantly influenced the total counts (p = 0.0194). The
marginal R\^2 explains 25% of the total variance, while the conditional
R\^2 explains 90% of the variance.

```{r}

estuary <- read.csv("../data_raw/Estuaries.csv", header = T)

estuary.glm <- glmer(Total ~ Modification + (1|Estuary), family = poisson(link = "log"), estuary)

estuary.glm.random <- glmer(Total ~ Modification + (1|Estuary) + (1|Estuary:Site), family = poisson(link = "log"), estuary)
summary(estuary.glm.random)

anova(estuary.glm, estuary.glm.random, test = "LRT")

r2(estuary.glm)
r2(estuary.glm.random)

estuary_means <- emmeans(estuary.glm.random, ~ Modification, type = "response")
plot_data <- as.data.frame(estuary_means)

ggplot(estuary, aes(x = Modification, y = Total)) +
  geom_jitter(aes(color = Estuary), width = 0.2, alpha = 0.5) +
 geom_pointrange(data = plot_data, 
                  aes(y = rate, ymin = asymp.LCL, ymax = asymp.UCL),
                  color = "black", 
                  size = 1) +
  labs(y = "Total", x = "Estuary Type") +
  theme_minimal()

```

*Problem 3* Transform the counts of Hydroid to presence/absence data.
Then model this variable as a function of Mod, fit the same 2 models as
in the previous exercise, and select between both. Communication your
results (plot + description of main results).

Including Site as a random effect significantly improved model fit
(chi\^2 = 6.21, p = 0.013). The Modified/Pristine Estuary had a
significant effect on Hydroid occurrence (z = 2.08, p = 0.037). Pristine
sites had a higher probability of Hydroid presence compared to Modified
sites.

```{r}
estuary$Presence <- (estuary$Hydroid > 0) + 0

hydroid.glm <- glmer(Presence ~ Modification + (1|Estuary), 
                     family = binomial(link = "logit"), data = estuary)

hydroid.glm.random <- glmer(Presence ~ Modification + (1|Estuary) + 
                              (1|Estuary:Site), 
                            family = binomial(link = "logit"), data = estuary)

anova(hydroid.glm, hydroid.glm.random, test = "LRT")

summary(hydroid.glm.random)

hydroid_plot_data <- estuary %>%
  group_by(Modification) %>%
  summarize(
    proportion = mean(Presence),
    n = n(),
    se = sqrt((proportion * (1 - proportion)) / n)
  )

hydroid_em <- emmeans(hydroid.glm.random, ~ Modification, type = "response")
hydroid_estimates <- as.data.frame(hydroid_em)

ggplot(estuary, aes(x = Modification, y = Presence)) +
  
  geom_jitter(aes(color = Estuary), width = 0.2, height = 0.05, alpha = 0.5) +
  
  geom_pointrange(data = hydroid_estimates, aes(y = prob, ymin = asymp.LCL,
                                                ymax = asymp.UCL),
                  color = "black", size = 1) +
  
  labs(y = "Probability of Presence (%)", x = "Estuary Status") +
  theme_classic()

```

#Causal Inference

```{r}
library(dagitty)
library(ggdag)
```

*Problem 1* Researchers are interested in whether using mosquito nets
(variable net) decreases an individual’s risk of contracting malaria
(variable malaria_risk). They collected observational data in
mosquito_nets.csv.

a)  Fit a linear model using net as explanatory and malaria_risk as
    response.

```{r}
nets <- read.csv("../data_raw/mosquito_nets.csv")
nets$net <- as.factor(nets$net)
nets_lm <- lm(malaria_risk~net, data = nets)
summary(nets_lm)
```

b)  Plot the DAG depicted in the slides using dagitty and
    library(ggdag), and find the minimal adjustment set.

```{r}
dag <- dagitty("
dag {
income -> net
income -> malaria_risk
income -> health
net <- income
net <- temperature
net <- eligible
net <- health
net <- household
net -> malaria_risk
eligible <- income
eligible <- household
eligible -> net
health -> malaria_risk
resistance -> malaria_risk
temperature -> malaria_risk
}
")

ggdag(dag, text = TRUE, use_labels = "name") +
  theme_dag()

# Find minimal adjustment set
adjustmentSets(dag, exposure = "net", outcome = "malaria_risk")
```

c)  Based on the previous DAG and minimal adjustment set, estimate the
    causal effect of bed net usage on malaria risk

```{r}
model_causal <- lm(malaria_risk ~ health + income + temperature + net, data = nets)

summary(model_causal)
```

d)  Write the R code that you would use to analyze these data.

Please see above.

*Problem 2* Use data(keeley) in library(piecewiseSEM) to build a
structural equation model. This model should consist of abiotic
favorability (abiotic) as a linear function of distance to coast
(distance); fire severity (firesev) as a linear function of stand age
(age); cover as a linear function of fire severity, and plant species
richness (rich) as a linear function of distance, plot heterogeneity
(hetero), abiotic and cover.

a)  Plot the SEM and use Fisher’s test of d-separation to assess
    goodness-of-fit. Are the assumed causal relationships consistent
    with the observed data? What is the explanatory power for the
    response variable of interest (plant species richness)?

```{r}
library(piecewiseSEM)
library(semEff)
data(keeley)
keeley_psem <- psem(
  lm(abiotic ~ distance, keeley),
  lm(firesev ~ age, keeley),
  lm(cover ~ firesev, keeley),
  lm(rich ~ distance + hetero + abiotic + cover, keeley)
)
summary(keeley_psem)
rsquared(keeley_psem)

plot(keeley_psem, node_attrs = data.frame(shape = "rectangle", color = "black", fillcolor = "grey"),
layout = "tree")


```

Fisher's C = 30.5, p value = 0.105 P value is \> 0.05, indicating that
the assumed causal relationships are consistent with the observed data.
R-squared for richness is 0.54, indicating that the model explains 54%
of the variation in plant species richness.

b)  Use library(semEff) to calculate the direct, indirect and total
    effects on the response variable of interest (plant species
    richness). Make a forest plot of the standardized total effects.

```{r}
keeley_psem.eff <- semEff(keeley_psem, R = 500)
rich_eff <- getTotEffTable(keeley_psem.eff, response = "rich")
summary(keeley_psem.eff, response = "rich")

#Still giving errors so I used ggplot to make the forest plot
plot_data <- as.data.frame(rich_eff)
plot_data$Variable <- rownames(plot_data)

colnames(plot_data)

ggplot(plot_data, aes(x = effect, y = Variable)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci), height = 0.2) +
  geom_point(size = 3)

```

# Multivariate Statistics

```{r}
library(vegan)
```

# Multivariate statistics

*Problem 1* Use data(dune) and data(dune.env) from library(vegan) to
create a Euclidean distance matrix after standardization using the first
(A1), second (Moisture) and fifth (Manure) variables of dune.env. Then
create a Bray-Curtis distance matrix using dune. Perform a Mantel test
on both distance matrices.

```{r}
data("dune")
data("dune.env")
vare.env.sta<-decostand(dune, method="standardize")
vare.env.euc<-vegdist(vare.env.sta, method="euclidean")
vare.bio.bray<-vegdist(dune, method="bray")
mantel(vare.bio.bray, vare.env.euc)
```

*Problem 2* Use data(dune) to:

a)  Make dendrograms of the results of hierarchical clustering using the
    single, average and complete methods after creating a Bray-Curtis
    distance matrix.

```{r}
vare.bio.bray<-vegdist(dune, method="bray")
vare.bio.clust<-hclust(vare.bio.bray, method="average")

plot(vare.bio.clust)
```

b)  Perform k-means partitions from 2 to 6 partitions and select the
    best partition using the Calinski criterion. Visualize the results.

```{r}
vare.bio.k<-cascadeKM(dune, inf.gr=2, sup.gr=6)
plot(vare.bio.k, sortg=T)
```

*Problem 3* Use data(dune) to:

a)  Perform a nonmetric multidimensional scaling (NMDS) with data(dune)
    after calculating Bray-Curtis distances among sites. Plot the
    results including the stress value, the size of the points should be
    proportional to Shannon diversity, and the color should represent
    the variable Management in data(dune.env).

```{r}
vare.bio.nmds<-metaMDS(dune, distance="bray", k=2, trymax = 100)
```

```{r}
vare.bio.nmds$stress
```

```{r}
stressplot(vare.bio.nmds)
```

```{r}
plot(vare.bio.nmds, type="t", display="sites")
```

```{r}
#package GGalt has been archived and removed from the CRAN repository, used an ellipse instead

library(see)
library(ggplot2)

shannon_dune <- diversity(dune, index="shannon")

dune.sp <- cbind(as.data.frame(dune), 
                Axis01 = vare.bio.nmds$points[,1], 
                Axis02 = vare.bio.nmds$points[,2], 
                Diversity = shannon_dune, 
                Management = dune.env$Management)

ggplot(data=dune.sp, aes(x=Axis01, y=Axis02, color=Management)) +
  geom_point(aes(size=Diversity), alpha = 0.7) +
  stat_ellipse(aes(fill = Management), geom = "polygon", alpha = 0.2, show.legend = FALSE) +
  scale_size_continuous(range = c(2, 6)) +
  scale_color_see() +
  scale_fill_see() +
  annotate("text", x = -0.4, y = 0.45, 
           label = paste("Stress =", round(vare.bio.nmds$stress, 3))) +
  labs(x="Axis 1", y="Axis 2") +
  theme_classic(base_size = 14)
```

b)  Perform a permutational multivariate analysis of variance
    (PERMANOVA) with 9,999 permutations using the dune dataset after
    calculating Bray-Curtis distances and the first (A1), second
    (Moisture) and fifth (Manure) variables of dune.env as explanatory.
    Report which explanatory variables are significant and their
    explanatory power.

```{r}
perm_set <- dune.env[, c("A1", "Moisture", "Manure")]

adonis_dune <- adonis2(dune ~ A1 + Moisture + Manure, data = perm_set, permutations = 9999, method = "bray", by = "terms")

print(adonis_dune)
```

All 3 variables were significant. (p value \< 0.05). Moisture has the
highest explanatory power, explaining 27.6% of variability. Manure
explains 25.2% of the variation, and A1 explains 16.8% of variation.
Total, these 3 variables explain 69.6% of the variability between sites.
