---
title: "PhoebeTeskey_Linear_Non-linear_Models_ENVS567"
output: html_document
---


```{r}
library(ggplot2)
library(easystats)
library(DHARMa)
library(MASS)
library(lme4)
library(ecostats)
library(GGally)
library(car)

```



*Linear (and non-linear) Models*

*Question 1:* Data in growth.txt come from a farm-scale trial of animal diets. There are two factors: diet and supplement. Diet is a factor with three levels: barley, oats and wheat. Supplement is a factor with four levels: agrimore, control, supergain and supersupp. The response variable is weight gain after 6 weeks. Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

Question: What diet and supplement increases weight in mice after 6 weeks?

```{r}
#Exploratory analyses

growth <- read.table("../data_raw/growth.txt", header = T)
weight_gain_mice <- lm(gain ~ diet * supplement, data = growth)
plot(parameters(weight_gain_mice))

ggplot(data = growth, aes(x = diet, y = gain, color = supplement))+
  geom_boxplot()

```

```{r}
#Fit model

model3 <- lm(gain ~ diet * supplement, data = growth)
summary(model3)

```


```{r}
#Validate model

check_model(model3, check=c("linearity", "homogeneity", "normality"))
check_homogeneity(model3)
plot(check_homogeneity(model3))

```

```{r}
#Test hypothesis and assess explanatory power
anova(model3)

```


```{r}
#r^2
r2(model3)

```

Communication of results (plot + description of main results):

No significant interaction between diet and supplement. The effect of the supplements was consistent across diets. Type of diet was significant, (F2,36 = 83.52, p < 0.001) and the barley diet had a higher gain than barley or wheat. The supplement was also significant, (F3,36 = 17.8, p < 0.001). The model accounted for 82% of the variance in weight gain, (R2 = 0.82) with standard error at 1.31.

```{r}
model3.means <- estimate_means(model3)
amount.means <- estimate_means(model3, by = "diet")

ggplot(data = model3.means, aes(x = diet, y = Mean, color=supplement)) +
  geom_pointrange(aes(ymin = CI_low, ymax = CI_high), position=position_dodge(width=0.5), size=0.5) +
  geom_pointrange(data=amount.means, aes(ymin = CI_low, ymax = CI_high), position=position_dodge(width=0.5), size=1.5, color="black") +
  geom_jitter(data = growth, aes(x=diet, y=gain, color=supplement), alpha=0.6, size=1.5,
              position=position_jitterdodge(jitter.width = 0.2))+
  scale_color_manual(values = c("firebrick4", "dodgerblue3", "green", "orange"))+
  ylab("Weight (g)")+
  xlab(NULL)+
  theme_classic()
```

*Question #2:*
Respiratory rate (Y) is expected to depend on body mass (X) by the power law, Y=aX, where  is the scaling exponent. The data are available in respiratoryrate_bodymass.txt.

a)	Analyze the data by transforming the data and then fit a linear model.

```{r}
breathe <- read.table("../data_raw/respiratoryrate_bodymass.txt", header = T)
breathe$logBodyMass <- log(breathe$BodyMass)
breathe$logRespiration <- log(breathe$Respiration)

fit.log <- lm(logRespiration ~ logBodyMass, data = breathe)
summary(fit.log)

```

For every 1% increase in body mass, the respiratory rate increases by approximately 0.86%

```{r}
ggplot(data = breathe, aes(x=Respiration, y=BodyMass))+
  geom_point()+
  geom_smooth(method ="lm", se = T, color = "red")+
  scale_y_continuous(trans = "log")+
  theme_classic()
```


```{r}
ggplot(data = breathe, aes(x=Respiration, y=BodyMass))+
  geom_point()+
  geom_smooth(color= "red", method = "glm", method.args = list(family = gaussian(link = 'log')))+
  theme_classic()
```


*Question 3:* Use the diminish.txt data (xv is explanatory (independent), yv is response variable (dependent)) to perform a simple linear regression and a polynomial (second-degree) regression.

a)	Compare both models and decide which model seems to be better?

b)	Make a scatterplot and include both regression lines.


```{r}
diminish <- read.table("../data_raw/diminish.txt", header = T)

fit.lm <- lm(yv ~ xv, data=diminish)

fit.poly <- lm(yv ~ xv + I(xv^2), data=diminish)
summary(fit.poly)
```

```{r}
anova(fit.lm, fit.poly, test='LRT')
```

```{r}
compare_performance(fit.lm, fit.poly)
```

```{r}
ggplot(data=diminish, aes(x=xv, y=yv))+
  geom_point()+
  geom_smooth(method = "lm", se = T, color = "red")+
  geom_smooth(method = "lm", formula = y ~ poly(x,2), se = T)+
  theme_classic()
```


*Question 4* Use the ozone.txt data to model the ozone concentration as a linear additive (no interactions) function of wind speed, air temperature and the intensity of solar radiation.

a)	Make a pairwise scatterplot and assess the collinearity of the explanatory variables

```{r}
ozone <- read.table("../data_raw/ozone.txt", header = T)

ggpairs(ozone)
```


```{r}
ozone.lm <- lm(ozone ~ rad+temp+wind, data=ozone)
summary(ozone.lm)
```

```{r}
check_model(ozone.lm, check=c("linearity", "homogeneity", "normality"))
```

```{r}
check_collinearity(ozone.lm)
```


```{r}
plot(check_collinearity(ozone.lm))
```


```{r}
anova(ozone.lm)
```


```{r}
anova(lm(ozone ~ wind+rad+temp, data=ozone))
```


```{r}
anova(lm(ozone ~ wind+temp+rad, data=ozone))
```


```{r}
Anova(ozone.lm, type = 2)
```


```{r}
r2(ozone.lm)
```

We explored the linear relationships between Ozone, Radiation, Wind and Temperature. Temperature and Radiation variables were positively associated with higher ozone levels (p-values < 0.05). Overall, our multiple regression model explained 60% (adjusted R2) of the variability in Ozone levels.


```{r}
plot(parameters(ozone.lm))
```


```{r}
plot(parameters(ozone.lm, standardize = "refit"))
```


*Question 5:* Use data(CO2) and assess the effects of the variables Treatment (categorical) and conc (ambient concentration, numerical) on the response uptake. Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

Question: Does ambient concentration of CO2 and treatment type affect uptake of CO2 in plants?

```{r}
#Exploratory analyses

data(CO2)

ggplot(CO2, aes(x = uptake, y = conc, color = Treatment)) +
  geom_point()

```


```{r}
#Fit model

model4 <- lm(uptake ~ Treatment * conc, data = CO2)
summary(model4)

```


```{r}
#Validate model

check_model(model4, check=c("linearity", "homogeneity", "normality"))


```


```{r}
#Test hypothesis and assess explanatory power

anova(model4)

```


```{r}
Anova(model4)
```

```{r}
r2(model4)
```

Communicate Results:
There was not enough evidence of different relationships between concentration of CO2 and uptake of CO2 among treatment types (F1,80 = 0.4, p-value = 0.5297). The chilled treatment type have on average 5 less units of CO2 than the nonchilled treatment type (F1,80 = 74.1267, p-value < 0.001). There was a positive relationship between concentration of CO2 and uptake of CO2 (b = 0.0198, F1, 80 = 28.5535, p value < 0.001).  Overall, the model explained 31% of the variance in uptake of CO2.


```{r}
#Communicate results with a plot
ggplot(CO2, aes(y=uptake, x=conc, color=Treatment))+
  geom_point()+
  geom_smooth(method = "lm", se = T)+
  scale_color_manual(values = c("green", "pink"))+
  ylab("Uptake of CO2")+
  xlab("Concentration of CO2")+
  theme_classic()
```


*Question 6*: Use the jaws.txt data (age explanatory, bone response).

	Perform a non-linear regression assuming an asymptotic exponential relationship: y=a(1-e^(-cx))

	Perform a non-linear regression assuming a Michaelis-Menten model: y=ax/(1+bx)

	Estimate the percentage of variation explained by both models (comparing them with a null model with only a constant).

	Compare both models using AIC.

	Make a scatterplot of the data and include both regression lines.


```{r}
jaws <- read.table("../data_raw/jaws.txt", header = T)

jaws.asymp<-nls(bone~a*age/(b+age), start=list(a=100, b=0.1), data=jaws)
summary(jaws.asymp)
```


```{r}
summary(nls(bone ~ SSmicmen(age, a, b), data=jaws))
```


```{r}
rss<-13.9^2*51 # residual sum of squares

null<-lm(bone~1, data=jaws) # null model with only a constant
summary(null)
```


```{r}
tss<-31.01^2*52 # total sum of squares

100*(tss-rss)/tss
```


```{r}
fit_mm <- nls(bone ~ a * age / (1 + b * age), 
              data = jaws, 
              start = list(a = 20, b = 0.1))

aic_asymp <- AIC(jaws.asymp)
aic_mm <- AIC(fit_mm)

summary(aic_asymp)
summary(aic_mm)
```


```{r}


age_range <- data.frame(age = seq(min(jaws$age), max(jaws$age), length.out = 200))

age_range$asymp_pred <- predict(jaws.asymp, newdata = age_range)
age_range$mm_pred <- predict(fit_mm, newdata = age_range)

ggplot(jaws, aes(x = age, y = bone)) +
  geom_point(alpha = 0.6, color = "darkgrey") +
  # Add Asymptotic Exponential line
  geom_line(data = age_range, aes(y = asymp_pred, color = "Asymptotic Exponential"), size = 1) +
  # Add Michaelis-Menten line
  geom_line(data = age_range, aes(y = mm_pred, color = "Michaelis-Menten"), size = 1, linetype = "dashed") +
  # Formatting
  scale_color_manual(name = "Models", values = c("Asymptotic Exponential" = "red", "Michaelis-Menten" = "blue")) +
  labs(title = "Bone Growth vs. Age: Model Comparison",
       x = "Age",
       y = "Bone") +
  theme_minimal()
```



