---
title: "PhoebeTeskey_GLMs"
output: html_document
---

```{r}
library(ggplot2)
library(DHARMa)
library(MASS)
library(easystats)
library(lme4)


library(ecostats)
library(GGally)
library(car)
library(parameters)
library(performance)
library(modelbased)
```


GLMs

1. The data in stork.txt display the stress-induced corticosterone levels circulating in the blood of European white storks and their survival over the subsequent five years of study. Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

Question: Does corticosterone level predict survival of European white storks over a period of 5 years?

```{r}
#Exploratory analyses
stork <- read.table("../data_raw/stork.txt", header = T)

ggplot(stork, aes(x = Corticosterone, y = Survival))+
  geom_point()


```

Fit Model:

```{r}
stork.glm<-glm(Survival ~ Corticosterone, family=binomial(link="logit"), data = stork)
summary(stork.glm)
```

Validate Model:

```{r}
simulateResiduals(stork.glm, plot = T)
```

Test hypothesis and assess explanatory power:

```{r}
anova(stork.glm, test = "LRT")
```

```{r}
r2(stork.glm)
```

Communicate results:
Animals with lower corticosterone levels have a higher rate of survival (-0.07980, Chisq = 3.8382, p-value 0.0501). The model explained 10% of the variance.

```{r}
ggplot(aes(x=Corticosterone, y=Survival), data=stork)+
  geom_point(size = 2, alpha = 0.9)+
  geom_smooth(method=glm, method.args = list(family = "binomial"), se=T)+
  theme_classic()
```


2. The clusters.txt dataset contains the response variable Cancers (the number of reported cancer cases per clinic) and the explanatory variable Distance (the distance from a nuclear plant to the clinic in kilometers). Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

1.) Question: Do cancer cases occur more often closer to a nuclear power plant?

```{r}
#2.) Exploratory analyses

cancer <- read.table("../data_raw/clusters.txt", header = T)

ggplot(data = cancer, aes(x = Cancers))+
  geom_histogram(color = "white", fill = "black")

```

3.) Fit model

```{r}
cancer.glm <- glm(Cancers ~ Distance, data = cancer, family = poisson)
summary(cancer.glm)
```

Validate Model:

```{r}
simulateResiduals(cancer.glm, plot = T)
```

```{r}
testDispersion(cancer.glm, plot = T)
```

```{r}

cancer.nb <- glm.nb(Cancers ~ Distance, data=cancer)
summary(cancer.nb)
```

Test hypotheses and assess explanatory power:

```{r}
simulateResiduals(cancer.nb, plot = T)
```

```{r}
testDispersion(cancer.nb, plot = T)
```

```{r}
anova(cancer.nb, test = "LRT")
```

```{r}
r2(cancer.nb)
```

Communicate Results:

There is no significant relationship between number of cancer cases and distance from a nuclear power plant. (Chisq = 1.6741, p value 0.19). The model explained 2.7% of the variance. Based on these results, we cannot conclude that living closer to a power plant increases the number of reported cancer cases.


```{r}
ggplot(data = cancer, aes(x = Distance, y = Cancers)) +
  geom_point()+
  geom_smooth(method = "glm", method.args = list(family = "poisson"), color = "red") +
  labs(x =  "Distance from nuclear power plant", y = "Cancer cases count")+
  theme_classic()
```

3. Use data(cbpp) from library(lme4) to model the number of cases of cattle plague (incidence) as a function of period. Control for herd size in your model. Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

Question: Is the number of cases of cattle plague influenced by time period or herd size?

```{r}
#Exploratory analyses
plague <- cbpp


```




















