---
title: "PhoebeTeskey_GLMs"
output: html_document
---

```{r}
library(ggplot2)
library(DHARMa)
library(MASS)
library(easystats)
library(lme4)


library(ecostats)
library(GGally)
library(car)
library(parameters)
library(performance)
library(modelbased)
```


GLMs

1. The data in stork.txt display the stress-induced corticosterone levels circulating in the blood of European white storks and their survival over the subsequent five years of study. Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

Question: Does corticosterone level predict survival of European white storks over a period of 5 years?

```{r}
#Exploratory analyses
stork <- read.table("../data_raw/stork.txt", header = T)

ggplot(stork, aes(x = Corticosterone, y = Survival))+
  geom_point()


```

Fit Model:

```{r}
stork.glm<-glm(Survival ~ Corticosterone, family=binomial(link="logit"), data = stork)
summary(stork.glm)
```

Validate Model:

```{r}
simulateResiduals(stork.glm, plot = T)
```

Test hypothesis and assess explanatory power:

```{r}
anova(stork.glm, test = "LRT")
```

```{r}
r2(stork.glm)
```

Communicate results:
Animals with lower corticosterone levels have a higher rate of survival (-0.07980, Chisq = 3.8382, p-value 0.0501). The model explained 10% of the variance.

```{r}
ggplot(aes(x=Corticosterone, y=Survival), data=stork)+
  geom_point(size = 2, alpha = 0.9)+
  geom_smooth(method=glm, method.args = list(family = "binomial"), se=T)+
  theme_classic()
```


2. The clusters.txt dataset contains the response variable Cancers (the number of reported cancer cases per clinic) and the explanatory variable Distance (the distance from a nuclear plant to the clinic in kilometers). Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

1.) Question: Do cancer cases occur more often closer to a nuclear power plant?

```{r}
#2.) Exploratory analyses

cancer <- read.table("../data_raw/clusters.txt", header = T)

ggplot(data = cancer, aes(x = Cancers))+
  geom_histogram(color = "white", fill = "black")

```

3.) Fit model

```{r}
cancer.glm <- glm(Cancers ~ Distance, data = cancer, family = poisson)
summary(cancer.glm)
```

Validate Model:

```{r}
simulateResiduals(cancer.glm, plot = T)
```

```{r}
testDispersion(cancer.glm, plot = T)
```

```{r}

cancer.nb <- glm.nb(Cancers ~ Distance, data=cancer)
summary(cancer.nb)
```

Test hypotheses and assess explanatory power:

```{r}
simulateResiduals(cancer.nb, plot = T)
```

```{r}
testDispersion(cancer.nb, plot = T)
```

```{r}
anova(cancer.nb, test = "LRT")
```

```{r}
r2(cancer.nb)
```

Communicate Results:

There is no significant relationship between number of cancer cases and distance from a nuclear power plant. (Chisq = 1.6741, p value 0.19). The model explained 2.7% of the variance. Based on these results, we cannot conclude that living closer to a power plant increases the number of reported cancer cases.


```{r}
ggplot(data = cancer, aes(x = Distance, y = Cancers)) +
  geom_point()+
  geom_smooth(method = "glm", method.args = list(family = "poisson"), color = "red") +
  labs(x =  "Distance from nuclear power plant", y = "Cancer cases count")+
  theme_classic()
```

*Question 3:* Use data(cbpp) from library(lme4) to model the number of cases of cattle plague (incidence) as a function of period. Control for herd size in your model. Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

Question: Is the number of cases of cattle plague influenced by time period or herd size?

```{r}
#Exploratory analyses
plague <- cbpp

ggplot(plague, aes(x = incidence, y = size))+
  geom_point()

ggplot(plague, aes(x = incidence, y = period))+
  geom_point()

```

Fit Model
```{r}
plague_glmm <- glmer(cbind(incidence, size - incidence) ~ period + (1 | herd), 
                    data = plague, 
                    family = binomial)
summary(plague_glmm)
```

Validate Model
```{r}
simulateResiduals(plague_glmm, plot = T)
```

Test hypothesis and assess explanatory power:

```{r}
drop1(plague_glmm, test = "Chisq")
```

```{r}
r2(plague_glmm)
```

Communicate Results: The model revealed a significant primary effect of period on disease incidence (chisq = 25.61, df = 3, p < 0.001). All subsequent periods after baseline (Period 1) showed a marked reduction in the probability of infection.

```{r}
pred <- estimate_relation(plague_glmm) #modelbased
pred.farm <- estimate_relation(plague_glmm, include_random = T)

ggplot(pred.farm, aes(x=herd, y=Predicted, group=period))+
  geom_line(color="darkgrey", alpha=0.6)+
  geom_point(data = plague, aes(x=herd, y=size, color = period), size=0.9, alpha=0.6, position = position_jitter(width = 0, height = 0.05))+
  geom_line(data = pred, aes(x = period, y = Predicted, color=period), linewidth = 2)+
  geom_ribbon(data = pred, aes(x = period, ymin = CI_low, ymax = CI_high, fill=period), alpha = 0.2) +
  scale_color_manual(values=c("firebrick4", "dodgerblue3", "green", "pink"))+
  scale_fill_manual(values=c("firebrick4", "dodgerblue3", "green", "pink"))+
  facet_wrap(~period)+
  ylab("Herd size")+
  xlab("Incidence of plague")+
  theme_classic()
```


*Question 4:* Use insects.csv to model the number of insects (count) as an additive function (no interaction) of habitat (categorical) and flowers (numerical). Previous studies have reported that habitat quality tends to drive the detection of insects (some habitats have no insects regardless of flowers). Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).












