---
title: "PhoebeTeskey_CausalInference"
output: html_document
---

```{r}
library(dagitty)
library(ggdag)
install.packages("")


```


1. Researchers are interested in whether using mosquito nets (variable net) decreases an individual’s risk of contracting malaria (variable malaria_risk). They collected observational data in mosquito_nets.csv.

a)	Fit a linear model using net as explanatory and malaria_risk as response.

```{r}
nets <- read.csv("../data_raw/mosquito_nets.csv")
nets$net <- as.factor(nets$net)
nets_lm <- lm(malaria_risk~net, data = nets)
summary(nets_lm)
```

b)	Plot the DAG depicted in the slides using dagitty and library(ggdag), and find the minimal adjustment set.

```{r}
dag <- dagitty("
dag {
income -> net
income -> malaria_risk
income -> health
net <- income
net <- temperature
net <- eligible
net <- health
net <- household
net -> malaria_risk
eligible <- income
eligible <- household
eligible -> net
health -> malaria_risk
resistance -> malaria_risk
temperature -> malaria_risk
}
")

ggdag(dag, text = TRUE, use_labels = "name") +
  theme_dag()

# Find minimal adjustment set
adjustmentSets(dag, exposure = "net", outcome = "malaria_risk")
```


c)	Based on the previous DAG and minimal adjustment set, estimate the causal effect of bed net usage on malaria risk

```{r}
model_causal <- lm(malaria_risk ~ health + income + temperature + net, data = nets)

summary(model_causal)
```

d)	Write the R code that you would use to analyze these data.

Please see above.

2. Use data(keeley) in library(piecewiseSEM) to build a structural equation model. This model should consist of abiotic favorability (abiotic) as a linear function of distance to coast (distance); fire severity (firesev) as a linear function of stand age (age); cover as a linear function of fire severity, and plant species richness (rich) as a linear function of distance, plot heterogeneity (hetero), abiotic and cover.

```{r}

```



a)	Plot the SEM and use Fisher’s test of d-separation to assess goodness-of-fit. Are the assumed causal relationships consistent with the observed data? What is the explanatory power for the response variable of interest (plant species richness)?

```{r}
library(piecewiseSEM)
library(semEff)
data(keeley)
keeley_psem <- psem(
  lm(abiotic ~ distance, keeley),
  lm(firesev ~ age, keeley),
  lm(cover ~ firesev, keeley),
  lm(rich ~ distance + hetero + abiotic + cover, keeley)
)
summary(keeley_psem)
rsquared(keeley_psem)

plot(keeley_psem, node_attrs = data.frame(shape = "rectangle", color = "black", fillcolor = "grey"),
layout = "tree")


```


Fisher's C = 30.5, p value = 0.105
P value is > 0.05, indicating that the assumed causal relationships are consistent with the observed data.
R-squared for richness is 0.54, indicating that the model explains 54% of the variation in plant species richness.


b)	Use library(semEff) to calculate the direct, indirect and total effects on the response variable of interest (plant species richness). Make a forest plot of the standardized total effects.


```{r}
keeley_psem.boot<-bootEff(keeley_psem, R = 500, parallel = "multicore")
keeley_psem.eff<-semEff(keeley_psem.boot, data = keeley)
summary(keeley_psem.eff, response = "rich")

#this code will not run, tried multiple different things to fix and nothing worked
# keeley_std <- stdEff(keeley_psem.eff)
# 
# forestPlot(keeley_std, 
#            response = "rich",
#            what = "total")
# Error:
# Error in nobs.default(m) : no 'nobs' method is available
# Error in attr(x, "formula") %||% { : invalid formula

```





