---
title: "PhoebeTeskey_Linear_Non-linear_Models_ENVS567"
output: html_document
---


```{r}
library(DHARMa)
library(parameters)
library(MASS)
library(lme4)
library(performance)
library(modelbased)
library(car)
library(ecostats)
```



*Linear (and non-linear) Models*

1. Data in growth.txt come from a farm-scale trial of animal diets. There are two factors: diet and supplement. Diet is a factor with three levels: barley, oats and wheat. Supplement is a factor with four levels: agrimore, control, supergain and supersupp. The response variable is weight gain after 6 weeks. Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

Question: What diet and supplement increases weight in mice after 6 weeks?

```{r}
#Exploratory analyses

growth <- read.table("data_raw/growth.txt", header = T)
weight_gain_mice <- lm(gain ~ diet * supplement, data = growth)
plot(parameters(weight_gain_mice))

ggplot(data = growth, aes(x = diet, y = gain, color = supplement))+
  geom_boxplot()

```

```{r}
#Fit model

model3 <- lm(gain ~ diet * supplement, data = growth)
summary(model3)

```


```{r}
#Validate model

check_model(model3, check=c("linearity", "homogeneity", "normality"))
check_homogeneity(model3)
plot(check_homogeneity(model3))

```

```{r}
#Test hypothesis and assess explanatory power
anova(model3)

```


```{r}
#r^2
r2(model3)

```

Communication of results (plot + description of main results):

No significant interaction between diet and supplement. The effect of the supplements was consistent across diets. Type of diet was significant, (F2,36 = 83.52, p < 0.001) and the barley diet had a higher gain than barley or wheat. The supplement was also significant, (F3,36 = 17.8, p < 0.001). The model accounted for 82% of the variance in weight gain, (R2 = 0.82) with standard error at 1.31.

```{r}
model3.means <- estimate_means(model3)
amount.means <- estimate_means(model3, by = "diet")

ggplot(data = model3.means, aes(x = diet, y = Mean, color=supplement)) +
  geom_pointrange(aes(ymin = CI_low, ymax = CI_high), position=position_dodge(width=0.5), size=0.5) +
  geom_pointrange(data=amount.means, aes(ymin = CI_low, ymax = CI_high), position=position_dodge(width=0.5), size=1.5, color="black") +
  geom_jitter(data = growth, aes(x=diet, y=gain, color=supplement), alpha=0.6, size=1.5,
              position=position_jitterdodge(jitter.width = 0.2))+
  scale_color_manual(values = c("firebrick4", "dodgerblue3", "green", "orange"))+
  ylab("Weight (g)")+
  xlab(NULL)+
  theme_classic()
```

Question #2:
Respiratory rate (Y) is expected to depend on body mass (X) by the power law, Y=aX, where  is the scaling exponent. The data are available in respiratoryrate_bodymass.txt.

a)	Analyze the data by transforming the data and then fit a linear model.

```{r}
breathe <- read.table("data_raw/respiratoryrate_bodymass.txt")
```



GLMs

1. The data in stork.txt display the stress-induced corticosterone levels circulating in the blood of European white storks and their survival over the subsequent five years of study. Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

```{r}

```


2. The clusters.txt dataset contains the response variable Cancers (the number of reported cancer cases per clinic) and the explanatory variable Distance (the distance from a nuclear plant to the clinic in kilometers). Analyze the data following this structure: 1. Question, 2. Exploratory analyses, 3. Fit model, 4. Validate model, 5. Test hypothesis and assess explanatory power, and 6. Communication of results (plot + description of main results).

1.) Question: Do cancer cases occur more often closer to a nuclear power plant?

```{r}
#2.) Exploratory analyses

cancer <- read.table("data_raw/clusters.txt", header = T)

ggplot(data = cancer, aes(x = Cancers))+
  geom_histogram(color = "white", fill = "black")

```

3.) Fit model

```{r}
cancer.glm <- glm(Cancers ~ Distance, data = cancer, family = poisson)
summary(cancer.glm)
```

```{r}

simulateResiduals(cancer.glm, plot = T)
```

```{r}
testDispersion(cancer.glm, plot = T)
```

```{r}

cancer.nb <- glm.nb(Cancers ~ Distance, data=cancer)
summary(cancer.nb)
```

```{r}
simulateResiduals(cancer.nb, plot = T)
```

```{r}
testDispersion(cancer.nb, plot = T)
```

```{r}
anova(cancer.nb, test = "LRT")
```

```{r}
r2(cancer.nb)
```
There is no significant relationship between number of cancer cases and distance from a nuclear power plant. (Chisq = 1.6741, p value 0.19)

```{r}
ggplot(data = cancer, aes(x = Cancers, y = Distance)) +
  geom_jitter(data = cancer, aes(x=Cancers, y=Distance), alpha=0.5, width=0.2)+
  ylab("Distance from nuclear power plant")+
  xlab("Cancer cases count")+
  theme_classic()
```

Mixed models

1. Researchers at the University of Arizona want to assess the germination rate of saguaros using a factorial design, with 3 levels of soil type (remnant, cultivated and restored) and 2 levels of sterilization (yes or no). The same experimental design was deployed in 4 different greenhouses. Each of the unique treatments was replicated in 5 pots. 6 seeds were planted in each pot.

a)	How many fixed treatments (unique combinations) exist?

b)	What is the total number of pots?

c)	What is the total number of plants measured?

d)	Write the R code that you would use to analyze these data.

```{r}

```


2. Use Estuaries.csv to model the counts of invertebrate species (Total) between modified or pristine estuaries (Mod) including Estuary as random effects (only intercepts). Then fit the same model but now also including Site in the random effects. Select between both models parsimoniously. Communication your results (plot + description of main results).


```{r}

estuary <- read.csv("data_raw/Estuaries.csv", header = T)

estuary.glm <- glmer(Total ~ Modification + (1|Estuary), family = poisson(link = "log"), estuary)

estuary.glm.random <- glmer(Total ~ Modification + (1|Estuary) + (1|Estuary:Site), family = poisson(link = "log"), estuary)
summary(estuary.glm.random)

anova(estuary.glm, estuary.glm.random, test = "LRT")

r2(estuary.glm)
r2(estuary.glm.random)



```






















