---
title: "seeds2_PhoebeTeskey"
output: html_document
---

```{r}
library(tidyverse)
library(lme4)
library(performance)
library(easystats)
```

I see a few questions here.

Question 1: What treatments create the tallest plants across 2 different species?
Question 2: Does soil type or sterility influence germination of 2 different species of plants?
Question 3: Do different treatments increase the rate of a plant's growth over a fixed period of time?

This document will investigate the 2nd question.

```{r}
seeds <- read.csv("../data_raw/seeds.csv")
head(seeds)
```

Looks like we have plants in shared pots, will need to control for that. 6 plants to 1 pot. 
3 soil types; STP, CULT, REM.
Sterile or nonsterile
2 different species
Germination or non-germination
4 different measurements, on day 13, day 18, day 23 and day 28.

May need to group data by each plant, there are 4 measurements per plant.

The response variable being measured this time is Germination Y/N. Possible interactions are soil type, sterility, and plant species.
Fixed effect: pot
Random effects: sterility, plant species and soil type

We do NOT care about plant heights, yay!

```{r}
#clean_data
seeds_factor <- seeds %>% mutate(across(where(is.character), as.factor))
```



```{r}
#Explore data
ggplot(seeds_factor, aes(x = germin))+
geom_bar()+
  facet_wrap(~species)

ggplot(seeds_factor, aes(x = germin))+
geom_bar()+
  facet_wrap(~soil)

ggplot(seeds_factor, aes(x = germin))+
geom_bar()+
  facet_wrap(~sterile)
```
Looks like a lot of variability. C species had a lot of germination, L species had similar numbers of germinated and non-germinated.
STP soil had a lot of germination. REM and CULT were similar rates of germination and non-germination. 
Sterile condition had big changes in both germinated and non-germinated.







