---
title: "seeds_PhoebeTeskey"
output: html_document
---

```{r}
library(tidyverse)

```

Question: What treatment increases average growth across 2 different species of plants?


```{r}
seeds <- read.csv("../data_raw/seeds.csv")
head(seeds)
```

Looks like we have plants in shared pots, might need to control for that. 6 plants to 1 pot. 
3 soil types; STP, CULT, REM.
Sterile or nonsterile
2 different species
Germination or non-germination
4 different measurements, on day 13, day 18, day 23 and day 28.

May need to group data by each plant, there are 4 measurements per plant.
Variables we care about are height, pot, plant, soil, sterile, species.

```{r}
#clean the data
seeds_longer <- pivot_longer(seeds, cols = hgt13:hgt28, names_to = "day", values_to = "height")
seeds_clean <- na.omit(seeds_longer)
seeds_clean <- seeds_clean %>% mutate(across(where(is.character), as.factor))
avg_growth_seeds <- seeds_clean %>% 
  group_by(pot, plant, soil, sterile, species) %>% 
  summarise(avg_growth = mean(height)) %>% 
  ungroup()

```


```{r}
#Explore data
ggplot(avg_growth_seeds, aes(x = soil, y = avg_growth))+
geom_point()+
  facet_wrap(~species)

ggplot(avg_growth_seeds, aes(x = sterile, y = avg_growth))+
geom_point()+
  facet_wrap(~species)

```

Treatents do appear to have an effect. 
For species "C": avg_growth higher with sterile conditions, avg_growth increases with CULT, REM, STP soil in that order. 
For species "L": avg_growth slightly higher with sterile conditions, no clear effect if any soil outperforms the other.
We want to control for random effects from pot.

```{r}

```

